Parameter Recovery Validation

### 2025-06-13 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): fix process duration with stick-breaking temporal prior description
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

**Identifiability proof**: For any $\lambda \neq 1$:
$$\mathcal{L}(T_{\max}, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_{\max}, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$$

The fixed anchor $t^*_1 = 0$ cannot be scaled, and gene parameters are independent of $T_{\max}$, breaking the symmetry.

@@

**Pattern classification framework**: The three expression patterns are defined by constraints on the absolute temporal parameters:

- **Pre-activation**: $t^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$
- **Transient**: $0.0 < t^*_{0,\text{on}i} < 2.5$, $\delta^*_i < 2.0$, and $R_{\text{on},i} > 2.0$
- **Sustained**: $0.0 < t^*_{0,\text{on}i} < 1.5$, $\delta^*_i > 2.5$, and $R_{\text{on},i} > 2.0$

@@

@@

**Stick-breaking cell time coordinates:**

The cell times are sampled using a stick-breaking process that ensures proper ordering and eliminates scaling symmetry:

\begin{align}
    t^*_1 &= 0 \quad \text{(process start, fixed)} \\
    t^*_N &= T_{\max} \quad \text{(process end)} \\
    \xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
    t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{(recursive construction)}
\end{align}

This construction automatically ensures $0 = t^*_1 < t^*_2 < \cdots < t^*_N = T_{\max}$ and eliminates the need for clipping operations.

@@

2. **Sample stick-breaking variables and construct cell times**:
   \begin{align}
   {\xi_j}^{(s)} &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
   {t^*_1}^{(s)} &= 0 \\
   {t^*_N}^{(s)} &= {T_{\max}}^{(s)} \\
   {t^*_j}^{(s)} &= {t^*_{j-1}}^{(s)} + {\xi_{j-1}}^{(s)}({T_{\max}}^{(s)} - {t^*_{j-1}}^{(s)})
   \end{align}

>>>>

**Theorem (Elimination of Scaling Symmetry):**

For any scaling factor $\lambda \neq 1$, the likelihood functions are not equal:
$$\mathcal{L}(T_{\max}, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_{\max}, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$$

**Proof:**
1. **Fixed anchor constraint**: The scaled system requires $t^*_1 = \lambda \cdot 0 = 0$ and $t^*_N = \lambda T_{\max}$
2. **Process duration independence**: Gene parameters $t^*_{0,\text{on}i}, \delta^*_i$ are sampled independently of $T_{\max}$ and do not scale
3. **Asymmetric scaling**: Scaling $T_{\max} \rightarrow \lambda T_{\max}$ and $t^*_j \rightarrow \lambda t^*_j$ changes the relative positioning of gene events within the process timeline
4. **Likelihood dependence**: The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$, which change under asymmetric scaling

Therefore, the likelihood is not invariant under parameter scaling, eliminating the scaling symmetry. □

@@

**Pattern Classification with Absolute Parameters:**

Using the expected process duration $\mathbb{E}[T_{\max}] = 5.0$, the pattern constraints become:

- **Pre-activation**: $t^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$ (gene activated before process begins)
- **Transient**: $0.0 < t^*_{0,\text{on}i} < 2.5$ and $\delta^*_i < 2.0$ and $R_{\text{on},i} > 2.0$ (complete cycle within process)
- **Sustained**: $0.0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i > 2.5$ and $R_{\text{on},i} > 2.0$ (long-duration activation)

These absolute thresholds preserve the same biological pattern meanings as the original relative formulation.

@@

**Cross-System Biological Interpretation:**

The same $T_{\max} = 5.0$ represents different real-world durations depending on the characteristic splicing rate $\beta$:

- **Immediate-early response** ($\beta = 5 \text{ hr}^{-1}$): Process duration = $5.0/5 = 1$ hour
- **Cell cycle progression** ($\beta = 1 \text{ hr}^{-1}$): Process duration = $5.0/1 = 5$ hours
- **Developmental transition** ($\beta = 0.1 \text{ hr}^{-1}$): Process duration = $5.0/0.1 = 50$ hours

Gene timing parameters maintain consistent biological interpretation across scales.

@@

**Stick-Breaking Cell Time Construction:**

The cell times $\{t^*_j\}_{j=1}^N$ are constructed using a stick-breaking process with fixed anchor points:

\begin{align}
t^*_1 &= 0 \quad \text{(fixed process start)} \label{eq-anchor-start} \\
t^*_N &= T_{\max} \quad \text{(fixed process end)} \label{eq-anchor-end} \\
\xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \label{eq-stick-breaking-weights} \\
t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{for } j \in \{2, \ldots, N-1\} \label{eq-stick-breaking-construction}
\end{align}

This construction automatically ensures:
- **Ordered constraint**: $0 = t^*_1 < t^*_2 < \cdots < t^*_N = T_{\max}$
- **Fixed boundaries**: Process start and end are anchored
- **Efficient sampling**: $O(N)$ complexity with no constraint violations

@@

2. **Sample stick-breaking cell times**:
   \begin{align}
   {\xi_j}^{(s)} &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
   {t^*_1}^{(s)} &= 0, \quad {t^*_N}^{(s)} = {T_{\max}}^{(s)} \\
   {t^*_j}^{(s)} &= {t^*_{j-1}}^{(s)} + {\xi_{j-1}}^{(s)}({T_{\max}}^{(s)} - {t^*_{j-1}}^{(s)}) \quad \text{for } j \in \{2, \ldots, N-1\}
   \end{align}
